---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
library(tidyverse)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# rclimateca

[![](http://cranlogs.r-pkg.org/badges/rclimateca)](https://cran.r-project.org/package=rclimateca) [![Travis-CI Build Status](https://travis-ci.org/paleolimbot/rclimateca.svg?branch=master)](https://travis-ci.org/paleolimbot/rclimateca) [![Coverage Status](https://img.shields.io/codecov/c/github/paleolimbot/rclimateca/master.svg)](https://codecov.io/github/paleolimbot/rclimateca?branch=master)

This package is designed to be an R interface to various Environment Canada datasets, including the [historical climate data archive](http://climate.weather.gc.ca/). In the future, the package may also provide access to [hydrometric data](https://wateroffice.ec.gc.ca/) using the same interface (currently in the form of the [hydatr package](https://github.com/paleolimbot/hydatr).

## Installation

You can install rclimateca from CRAN with:

```{r cran-install, eval = FALSE}
install.packages("rclimateca")
```

Or the development version from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("paleolimbot/rclimateca")
```

If you can load the package, everything worked!

```{r}
library(rclimateca)
```

## Example

The workflow for which this **rclimateca** is designed is as follows:

1. Find climate stations using `ec_climate_search_locations()` or `ec_climate_geosearch_locations()`.
2. Download the data using `ec_climate_data()` or `ec_climate_mudata()`.

For example, to find recent daily data for a location close to Wolfville, Nova Scotia, one might do the following::

```{r}
ec_climate_geosearch_locations(
  "Wolfville NS", timeframe = "daily",
  year = 2017, limit = 5
)
```

```{r}
ec_climate_data(
  "WHITE ROCK NS 6507", timeframe = "daily",
  start = "2017-01-01", end = "2017-12-31"
)
```

## More information

See `vignette("ec_climate", package = "rclimateca")` and `?ec_climate_data` for more information about using **rclimateca** with the Environment Canada historical climate archive.
