---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# rclimateca

[![](http://cranlogs.r-pkg.org/badges/rclimateca)](https://cran.r-project.org/package=rclimateca) [![Travis-CI Build Status](https://travis-ci.org/paleolimbot/rclimateca.svg?branch=master)](https://travis-ci.org/paleolimbot/rclimateca)

Fetching data from Environment Canada's archive has always been a bit of a chore. In the old days, it was necessary to download data one click at a time from the [organization's search page](http://climate.weather.gc.ca/historical_data/search_historic_data_e.html). To bulk download hourly data would require a lot of clicks and a good chance of making a mistake and having to start all over again. There are several R solutions online (posted by [Headwater Analytics](http://www.headwateranalytics.com/blog/web-scraping-environment-canada-weather-data) and [From the Bottom of the Heap](http://www.fromthebottomoftheheap.net/2015/01/14/harvesting-canadian-climate-data/) ), but both solutions are mostly single-purpose, and don't solve the additional problem of trying to find climate locations near you. In the [rclimateca package](https://cran.r-project.org/package=rclimateca), I attempt to solve both of these problems to produce filtered, plot-ready data from a single command.

## Installation

You can install rclimateca from CRAN with:

```{r cran-install, eval = FALSE}
install.packages("rclimateca")
```

Or the development version from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("paleolimbot/rclimateca")
```

## Finding climate stations

We will start with finding sites near where you're interested in. Sometimes you will have a latitude and longitude, but most times you will have a town or address. Using the [prettymapr](https://cran.r-project.org/package=prettymapr) packages 'geocode' function, the `getClimateSites()` function looks up locations near you.

```{r}
library(rclimateca)
getClimateSites("gatineau QC")
```

If you also need data for a set of years, you can also pass a vector of years to further refine your data.

```{r}
getClimateSites("gatineau QC", year=2014:2016)
```

If you need to access these columns programmatically somehow, you should pass `nicenames=TRUE`.

```{r}
sites <- getClimateSites("gatineau QC", year=2014:2016, nicenames=TRUE)
names(sites)
```

If you'd like to apply your own subsetting operation, the entire dataset is also available through this package.

```{r}
data("ecclimatelocs")
names(ecclimatelocs)
```

## Downloading data

Downloading data is accomplished using the `getClimateData()` function, or if you'd like something less fancy, the `getClimateDataRaw()` function. There is documentation in the package for both, but `getClimateData()` has all the bells and whistles, so I will go over its usage first. You will first need a `stationID` (or a vector of them) - in our case I'll use the one for Chelsea, QC, because I like [the ice cream there](http://www.lacigaleicecream.ca/).

```{r}
df <- getClimateData(5585, timeframe="daily", year=2015)
str(df)
```

Boom! Data! For programmatic access to columns, it is usually adventageous to pass `nicenames=TRUE`.

```{r}
df <- getClimateData(5585, timeframe="daily", year=2015, nicenames=TRUE)
names(df)
```

The package can also melt the data for you (à la [tidyr](https://cran.r-project.org/package=tidyr)) so that you can easily use [ggplot](https://cran.r-project.org/package=ggplot2) to visualize.

```{r, warning=FALSE, message=FALSE, results='hide'}
library(ggplot2)
df <- getClimateData(5585, timeframe="daily", year=2015, format="long")
ggplot(df, aes(parsedDate, value)) + geom_line() + 
  facet_wrap(~param, scales="free_y")
```

The function can accept a vector for most of the parameters, which it uses to either download multiple files or to trim the output, depending on the parameter. How to Chelsea, QC and Kentville, NS stack up during the month of November (Pretty similar, as it turns out...)?

```{r, warning=FALSE, results='hide'}
df <- getClimateData(c(5585, 27141), timeframe="daily", year=2015, month=11, format="long")
ggplot(df, aes(parsedDate, value, col=factor(stationID))) + 
  geom_line() + facet_wrap(~param, scales="free_y")
```

You will also notice that a little folder called `ec.cache` has popped up in your working directory, which contains the cached files that were downloaded from the Environment Canada site. You can disable this by passing `cache=NULL`, but I don't suggest it, since the cache will speed up running the code again (not to mention saving Environment Canada's servers) should you make a mistake the first time.

This function can download a whole lot of data, so it's worth doing a little math for yourself before you overwhelm your computer with data that it can't all load into memory. As an example, I tested this function by downloading daily data for every station in Nova Scotia between 1900 and 2016, which took 2 hours, nearly crashed my computer, and resulted in a 1.3 **gigabyte** data.frame. You can do a few things (like ensure `checkdate=TRUE` and, if you're using `format="long"`, `rm.na=T`) to make your output a little smaller, or you can pass `ply=plyr::a_ply` to just cache the files so you only have to download them the once.

## Using with MUData

The rclimateca package can also output data in [MUdata format](http://github.com/paleolimbot/mudata), which includes both location data and climate data in an easily plottable object.

```{r, results='hide'}
library(mudata2)
md <- getClimateMUData(c(27141, 6354), year=1999, month=7:8, timeframe="daily")
autoplot(md)
```

## A little on how it works

The code behind this package is [available on GitHub](http://github.com/paleolimbot/rclimateca), but it is fairly extensive and designed to tackle all of the corner cases that make writing a package so much more difficult than a script that runs once. Essentially, it's very close to the solution posted on [From the Bottom of the Heap](http://www.fromthebottomoftheheap.net/2015/01/14/harvesting-canadian-climate-data/) and in the [documentation itself](ftp://client_climate@ftp.tor.ec.gc.ca/Pub/Get_More_Data_Plus_de_donnees/Readme.txt). From the documentation:

```
for year in `seq 1998 2008`;do for month in `seq 1 12`;do wget --content-disposition "http://climate.weather.gc.ca/climate_data/bulk_data_e.html?format=csv&stationID=1706&Year=${year}&Month=${month}&Day=14&timeframe=1&submit= Download+Data" ;done;done

WHERE; 
 year = change values in command line (`seq 1998 2008)
 month = change values in command line (`seq 1 12)
 format= [csv|xml]: the format output
 timeframe = 1: for hourly data 
 timeframe = 2: for daily data 
 timeframe = 3 for monthly data 
 Day: the value of the "day" variable is not used and can be an arbitrary value 
 For another station, change the value of the variable stationID
 For the data in XML format, change the value of the variable format to xml in the URL. 
```

This is, of course, the same as the `genURLS()` solution except for bash and wget instead of for R. This package uses the "format=csv" option, which produces a somewhat malformed CSV (there is quite a bit of header information). The "guts" of the read operation are as follows:

```{r, warning=FALSE}
# download the file
library(httr)
url <- "http://climate.weather.gc.ca/climate_data/bulk_data_e.html?format=csv&stationID=1706&Year=2008&Month=5&Day=14&timeframe=1&submit=Download+Data"
connect <- GET(url)
x <- content(connect, as="text", encoding="UTF-8")

# find the second emtpy line and start the read.csv after that line
xlines <- readLines(textConnection(x))
empty <- which(nchar(xlines) == 0)
empty <- empty[empty != length(xlines)]
# read the data frame
df <- read.csv(textConnection(x), 
               skip=empty[length(empty)], 
               stringsAsFactors = F, check.names = F)
```

I am sure is cringing as I read the entirity of each CSV file into memory, but they should be reminded that (a) R is terrible for memory usage anyway, and (b) the files are already "chunked up" by Environment Canada so that there is no chance of getting an unwieldy-sized file by accident.

## Wrapping it up

That's it! Hopefully now you can all download unlimited quantities of data in pure bliss (as long as Environment Canada keeps its URLs consistent).
